Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning

### 2022-05-14 manuscript/manuscript.Rmd
More touching up

In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting transition events.
We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
When machine learning models are able to observe transition phenomena, as in the stochastic case, they generally outperformed MCMC-based forecasting but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.

@@

Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
This paper is not intended to give a complete answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve; instead, this study aims to be as an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of forecasting problems.

>>>>

In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting transition events.
We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
When machine learning models are able to observe transition phenomena, as in the stochastic case, they generally outperformed MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
Yet, there are many nuances to these assessments as examining summary statistics, CRPS and log probability scores obscure finer detailed components of the forecasts, like the pronounced bimodality of ML forecasts for the stochastic transition case.

@@

Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
This paper is not intended to give a complete answer to whether machine learning models are the the best methods to use on any tipping point forecasting problem as this will take numbers of studies to resolve; instead, this study aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of forecasting problems.
Since we have shown that ML models can make good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are promising methods for critical transition forecasting problems.
